Results

### 2016-02-08 doc/paper/results.Rmd
updated results text

In addition to DeepSNV, we analyzed our $10^5$ genomes/$\mu$l influenza populations with the variant caller Lofreq. Lofreq had marginally decreased sensitivity compared DeepSNV when applied to the variant frequencies $\geq$ 1.0% but marginally increased sensitivity when applied to variant frequencies $\l$ 1.0% (`r lofreq`). Additionally, Lofreq had specificity near that seen in our cell culture benchmarking experiment, and much better than DeepSNV prior to our quality controls. This is not completely surprising as the Lofreq algorithm takes MapQ and Phred scores into account when calling variants, something DeepSNV does not do. Additionally, Lofreq was not prone to calling false positives near the ends of reads _(verify whether or not it uses read position)_. _Show distributions of Lofreq variants?_ The "out of the box" accuracy of Lofreq was better than that of DeepSNV. However, our modified DeepSNV protocol greatly surpasses that of Lofreq, and as was the case with DeepSNV, the Lofreq ROCs (`r lofreq` A) suggests that more stringent cut offs will greatly sacrifice sensitivity for specificity.

##### Buyer beware

#### Lower populations

Influenza patient samples vary in input levels by several orders of magnitude depending on a variety of factors ranging from biological such as day post infection to methodological such as how much mucus is collected on the swab. To ensure accuracy across a range of input levels we diluted our known mutant mixes serially into viral media. We found both decreases in sensitivity and specificity accompanied lower input samples (`r lower`). The decreased sensitivity is expected as the likelihood of detecting rare events decreases with sample size. The decreased specificity is most likely due to a greater dependence on RT-PCR to amplify our signal to detectable limits. Is it not surprising that an amplification of noise accompanies this dependence. However, our results do highlight the importance of controlling for input levels when comparing populations of RNA viruses.

If RT-PCR errors are responsible for the decreased specificity found at lower input levels we would expect such errors to be controlled by processing the samples in duplicate. To test this hypothesis, a subset of the $10^4$ genomes/$\mu$l were run in duplicate. The 0.2% was not run as we expected a decrease in sensitivity would accompany this approach and had little hope of finding these rare variants. Two RT-PCR reactions were run from a new RNA prep. These samples were processed separately and sequenced on the same lane of an Illumina Hiseq. The aforementioned cut offs were applied and for a variant call to be considered it needed to be found in both duplicates. This drastically improved our specificity and brought it to a comparable level as the $10^5$ samples (`r dups`). Unexpectedly, an increase in sensitivity accompanied this sequencing run. This is most likely do to a cleaner library prep with less sample loss during size selection.

>>>>

In addition to DeepSNV, we analyzed our $10^5$ genomes/$\mu$l influenza populations with the variant caller Lofreq. Lofreq had marginally decreased sensitivity compared DeepSNV when applied to the variant frequencies $\geq$ 1.0% but marginally increased sensitivity when applied to variant frequencies $\l$ 1.0% (`r lofreq`). Additionally, Lofreq had specificity near that seen in our cell culture benchmarking experiment, and much better than DeepSNV prior to our quality controls. This is not completely surprising as the Lofreq algorithm takes MapQ and Phred scores into account when calling variants, something DeepSNV does not do. Additionally, Lofreq was not prone to calling false positives near the ends of reads _(verify whether or not it uses read position)_. _Show distributions of Lofreq variants?_ The "out of the box" accuracy of Lofreq was better than that of DeepSNV. However, our modified DeepSNV protocol greatly surpasses that of Lofreq.  `r Lofreq` A suggests that more stringent cut offs will greatly sacrifice sensitivity for specificity.

Lofreq has been reported to have perfect specificity and a sensitivity of 96% for variants present at 0.2%. However, this accuracy reflects a best case scenario using simulated sequencing reads. As was the case with DeepSNV, the accuracy of Lofreq was much lower when applied under conditions that often face virologist using patient derived samples. Our data suggests that is inappropriate to apply Lofreq or DeepSNV under diverse experimental conditions and expect the performance to be unaffected.

It may be possible to further filter the putative variants called by Lofreq and achieve a similar accuracy to what was achieved using DeepSNV. However, because DeepSNV was much less computationally intensive, more user friendly _not sure if I can or should say that_, and capable of high accuracy, we continued to benchmark DeepSNV using the less concentrated samples from our dilution series.

_Include Lofreq in supplimentary?_

#### Lower input levels

Influenza patient samples vary in input concentration by several orders of magnitude depending on a variety of factors ranging from biological such as day post infection to methodological such as how much mucus is collected on the swab. To ensure accuracy across a range of input levels we diluted our known mutant mixes serially into viral media. We found both decreases in sensitivity and specificity accompanied lower input samples (`r lower`). The decreased sensitivity is expected as the likelihood of detecting rare events decreases with sample size. The decreased specificity is most likely due to a greater dependence on RT-PCR to amplify our signal to detectable limits. Is it not surprising that an amplification of noise accompanies this dependence. However, our results do highlight the importance of controlling for input levels when comparing populations of RNA viruses.

If RT-PCR errors are responsible for the decreased specificity found at lower input levels we would expect such errors to be controlled by processing the samples in duplicate. To test this hypothesis, a subset of the $10^4$ genomes/$\mu$l were run in duplicate. The 0.2% was not run as we expected a decrease in sensitivity would accompany this approach and had little hope of finding these rare variants based on the sensitivity we saw in `r lower` B. Two RT-PCR reactions were run from a new RNA prep. These samples were processed separately and sequenced on the same lane of an Illumina Hiseq. The aforementioned cut offs were applied and for a variant call to be considered it needed to be found in both duplicates. This drastically improved our specificity and brought it to a comparable level as the $10^5$ samples (`r dups`). Unexpectedly, an increase in sensitivity accompanied this sequencing run. This is most likely do to a superior library prep with less sample loss during size selection.
